---
title: "Getting started with tweedieDistr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with tweedieDistr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6,
  fig.height = 4,
  out.width = "90%",
  fig.align  = "center"
)
```

```{r setup}
library(tweedieDistr)
library(ggplot2)
library(ggdist)
```

## Introduction

The **Tweedie distribution** with power parameter $p \in (1, 2)$ is a member
of the exponential dispersion model family that naturally mixes a point mass at
zero with a continuous positive component. It is particularly useful for
**intermittent demand** data, insurance loss modelling, and any other setting
where exact zeros co-exist with strictly positive observations.

`tweedieDistr` implements the distribution in two complementary styles:

1. **Base-R `stats` interface** — `dtweedie()`, `ptweedie()`, `qtweedie()`,
   `rtweedie()` — following the same conventions as `dnorm()`, `pnorm()`, etc.
2. **`distributional` interface** — `dist_tweedie()` — returning a fully
   featured distributional object that works seamlessly with tidy forecasting
   packages such as [`fable`](https://fable.tidyverts.org/) and the
   [`ggdist`](https://mjskay.github.io/ggdist/) plotting layer.

---

## The Tweedie distribution

### Compound Poisson--Gamma representation

A Tweedie variable $X$ with parameters $(\mu, \phi, p)$ can be written as a
compound sum

$$X = \sum_{i=1}^{N} G_i, \qquad N \sim \mathrm{Poisson}(\lambda),
\quad G_i \overset{\text{iid}}{\sim} \mathrm{Gamma}(\alpha, \beta),$$

with

$$\lambda = \frac{\mu^{2-p}}{\phi(2-p)}, \qquad
\alpha = \frac{2-p}{p-1}, \qquad
\beta = \frac{1}{\phi(p-1)\mu^{p-1}}.$$

When $N = 0$ the sum is defined as $X = 0$, so the distribution has a point
mass at zero with probability

$$P(X = 0) = e^{-\lambda}.$$

For $x > 0$ the density is continuous and is evaluated via the series expansion
of Dunn & Smyth (2005), implemented in C++ through Rcpp and RcppArmadillo.

### Moments

| Quantity | Formula |
|---|---|
| Mean | $\mu$ |
| Variance | $\phi\mu^p$ |
| Skewness | $p\sqrt{\phi}\,\mu^{p/2 - 1}$ |
| Excess kurtosis | $p(2p-1)\phi\,\mu^{p-2}$ |

---

## The `stats`-style interface

### Density — `dtweedie()`

```{r density-basic}
d <- dist_tweedie(mean = 3, dispersion = 0.8, power = 1.5)

tibble::tibble(dist = d) |>
  ggplot(aes(xdist = dist)) +
  stat_slab(fill = "#2171b5", alpha = 0.75, n = 1001) +
  labs(x = "x", y = "Density",
       title = expression(paste("Tweedie density  (",
                                mu, "=3, ", phi, "=0.8, p=1.5)"))) +
  theme_minimal()
```

The spike at $x = 0$ reflects the point mass $P(X = 0) = e^{-\lambda}$; the
continuous slab for $x > 0$ is the Gamma mixture component.

The `log` argument returns log-densities, which is numerically more stable for
likelihood computations:

```{r density-log}
dtweedie(c(0, 1, 2, 5), mean = 3, dispersion = 0.8, power = 1.5, log = TRUE)
```

### Effect of the power parameter

```{r density-power, fig.cap="Tweedie density across power values."}
powers <- c(1.2, 1.5, 1.8)

tibble::tibble(
  power = factor(paste0("p = ", powers), levels = paste0("p = ", rev(powers))),
  dist  = dist_tweedie(mean = 3, dispersion = 0.8, power = powers)
) |>
  ggplot(aes(xdist = dist, y = power, fill = power)) +
  stat_slab(alpha = 0.8, n = 1001) +
  scale_fill_manual(values = c("#e41a1c", "#2171b5", "#4daf4a")) +
  labs(x = "x", y = NULL, fill = NULL,
       title = expression(paste("Effect of p  (", mu, "=3, ", phi, "=0.8)"))) +
  theme_minimal() +
  theme(legend.position = "none")
```

Higher $p$ shifts more probability mass toward zero and produces a heavier
right tail.

### Distribution function — `ptweedie()`

```{r cdf}
q_grid <- seq(0, 8, by = 0.5)
cdf    <- ptweedie(q_grid, mean = 3, dispersion = 0.8, power = 1.5)
cbind(q = q_grid, CDF = round(cdf, 4))
```

The standard `lower.tail` and `log.p` arguments are supported, e.g. to
compute $P(X > 4)$:

```{r cdf-uppertail}
ptweedie(4, mean = 3, dispersion = 0.8, power = 1.5, lower.tail = FALSE)
```

### Quantile function — `qtweedie()`

`qtweedie()` uses a hybrid Newton--Raphson / bisection algorithm.

```{r quantile}
probs <- c(0.10, 0.25, 0.50, 0.75, 0.90, 0.95, 0.99)
qtweedie(probs, mean = 3, dispersion = 0.8, power = 1.5)
```

The boundary cases are handled correctly:

```{r quantile-boundary}
qtweedie(c(0, 1), mean = 3, dispersion = 0.8, power = 1.5)
```

### Random generation — `rtweedie()`

`rtweedie()` uses the exact compound Poisson--Gamma representation.

```{r rtweedie}
set.seed(2025)
samples <- rtweedie(500, mean = 3, dispersion = 0.8, power = 1.5)

ggplot() +
  geom_histogram(aes(x = samples, y = after_stat(density)),
                 bins = 30, fill = "#bdd7e7", colour = "white") +
  geom_function(fun = \(x) dtweedie(x, mean = 3, dispersion = 0.8, power = 1.5),
                xlim = c(1e-4, max(samples)),
                colour = "#2171b5", linewidth = 1) +
  labs(x = "x", y = "Density",
       title = "rtweedie() samples vs theoretical density") +
  theme_minimal()
```

### Vectorised arguments

All four functions recycle their arguments in the same way as base R:

```{r vectorised-dpqr}
# density for three different means at x = 2
dtweedie(2, mean = c(1, 3, 5), dispersion = 0.8, power = 1.5)

# quantile for two distributions simultaneously
qtweedie(0.9, mean = c(2, 5), dispersion = c(0.5, 1.0), power = 1.5)
```

---

## The `distributional` interface

`dist_tweedie()` wraps the distribution parameters in a
[`distributional`](https://github.com/mitchelloharawild/distributional) object.
This enables a unified, pipe-friendly API and integrates directly with
[`ggdist`](https://mjskay.github.io/ggdist/) for plotting, as well as packages
like `fable` that use `distributional` objects to represent forecast
distributions.

### Creating a distribution object

```{r dist-create}
d <- dist_tweedie(mean = 3, dispersion = 0.8, power = 1.5)
d
```

### Moments

```{r dist-moments}
mean(d)
distributional::variance(d)
distributional::skewness(d)
distributional::kurtosis(d)
```

### Density, CDF, and quantiles

```{r dist-dpq}
density(d, at = c(0, 1, 2, 3, 5))
distributional::cdf(d, q = c(1, 3, 5))
quantile(d, p = c(0.25, 0.5, 0.75, 0.9))
```

### Random generation

```{r dist-generate}
set.seed(2025)
distributional::generate(d, times = 10)
```

### Support

```{r dist-support}
distributional::support(d)
```

### Plotting with `ggdist`

`dist_tweedie()` objects integrate directly with `ggdist` via the `xdist`
aesthetic. `stat_halfeye()` shows the density slab together with a median point
and 66 %/95 % credible intervals.

```{r dist-halfeye}
tibble::tibble(dist = d) |>
  ggplot(aes(xdist = dist)) +
  stat_halfeye(fill = "#2171b5", alpha = 0.75, n = 1001) +
  labs(x = "x", y = NULL,
       title = expression(paste("dist_tweedie  (",
                                mu, "=3, ", phi, "=0.8, p=1.5)"))) +
  theme_minimal()
```

### Vectorised distribution objects

`dist_tweedie()` accepts vectors of parameters, returning one distribution
object per row. This is useful when each observation in a dataset has different
fitted parameters — and it pairs naturally with `ggdist` ridge plots.

```{r dist-vec}
d_vec <- dist_tweedie(
  mean       = c(1, 3, 5),
  dispersion = c(0.5, 0.8, 1.2),
  power      = c(1.3, 1.5, 1.8)
)
d_vec

mean(d_vec)
quantile(d_vec, p = 0.9)
distributional::variance(d_vec)
```

```{r dist-vec-plot, fig.cap="Ridge plot of three dist_tweedie distributions."}
labs <- c("(1, 0.5, 1.3)", "(3, 0.8, 1.5)", "(5, 1.2, 1.8)")
tibble::tibble(
  label = factor(labs, levels = rev(labs)),
  dist  = d_vec
) |>
  ggplot(aes(xdist = dist, y = label, fill = label)) +
  stat_slab(alpha = 0.8, n = 1001) +
  scale_fill_manual(values = c("#e41a1c", "#2171b5", "#4daf4a")) +
  labs(x = "x", y = expression(paste("(", mu, ", ", phi, ", p)")), fill = NULL,
       title = "Vectorised dist_tweedie distributions") +
  theme_minimal() +
  theme(legend.position = "none")
```

---

## Numerical accuracy

The density is computed via the weighted series expansion of Dunn & Smyth
(2005) in a numerically stable form using the log-sum-exp trick. The
implementation is validated against the `tweedie` package (Dunn, 2023), which
provides an independent reference implementation.

```{r accuracy, eval = requireNamespace("tweedie", quietly = TRUE)}
x_test <- c(0, 0.5, 1, 2, 5)
our    <- dtweedie(x_test, mean = 2, dispersion = 0.8, power = 1.5)
ref    <- tweedie::dtweedie(x_test, mu = 2, phi = 0.8, power = 1.5)

cbind(x = x_test, tweedieDistr = round(our, 8), tweedie_pkg = round(ref, 8),
      abs_diff = format(abs(our - ref), scientific = TRUE, digits = 3))
```

---

## References

Dunn, P. K. (2023). *tweedie: Tweedie Exponential Family Models*. R package
version 2.3.5. <https://CRAN.R-project.org/package=tweedie>.

Dunn, P. K., & Smyth, G. K. (2005). Series evaluation of Tweedie exponential
dispersion model densities. *Statistics and Computing*, 15(4), 267--280.
<https://doi.org/10.1007/s11222-005-4070-y>.

Jørgensen, B. (1987). Exponential dispersion models. *Journal of the Royal
Statistical Society, Series B*, 49(2), 127--162.
<https://doi.org/10.1111/j.2517-6161.1987.tb01685.x>.
